---
title: "autoimage README"
output: github_document
---

[![Build Status](https://travis-ci.org/jpfrench81/autoimage.svg?branch=master)](https://travis-ci.org/jpfrench81/autoimage)

## autoimage

The goal of the **autoimage** package is to easily plot a sequence of 
images with corresponding color scales, i.e., a sequence of heatmaps, 
with straightforward, native options for projection of geographical 
coordinates.

The package makes it simple to add lines and points
to these image plots, even when the coordinates are projected. 

The package allows for seamless plotting of gridded 
data on a non-regular grid or even data that is not on a grid.

```{r setup, include=FALSE, render=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(autoimage)
```

## Installation

The release version of **autoimage** can be installed in R using the command `install.packages("autoimage")`.

The development version of `autoimage` can be installed in R using the command  
```
# install.packages("devtools")
devtools::install_github("jpfrench81/autoimage")
```

## Examples

The most important functions are the `pimage` and `autoimage` functions.
`pimage` creates a heat map with a color scale for data on
a regular or irregular grid.  It can also create an image for
data that is not on a grid using interpolation.

The package also makes it easy to add lines and points to these images, 
with several common world maps being provided automatically through
the `maps` package.

We illustrate some basic usage of the `pimage` function.

We begin by display images for data on an irregular grid, including
national boundaries and legends.

```{r}
# display some narccap data on an irregular grid
data(narccap)
# heat map for data on an irregular grid with national boundaries
pimage(lon, lat, tasmax[,,1], map = 'world', legend = "horizontal")
# same plot but with projection
pimage(lon, lat, tasmax[,,1], map = 'world', legend = "vertical",
       proj = 'bonne', parameters = 45)
```

We next display an image for data that is not on a grid.  Specifically,
we create image plots for Colorado geochemical measurements.  We 
also include the Colorado border in the image and the locations
of the observed coordinates.  Note that the `lines.args` and `points.args`
arguments are used to customize the appearance of the lines and points.

```{r}
# colorado data not on a grid
# show observed locations on image,
# along with Colorado border
data(co, package = 'gear') # load colorado geochemical measurements
data(copoly) # load colorado borders
# create list with observed points
copoints <- list(x = co$lon, y = co$lat) 
reset.par() # reset graphics device
pimage(co$longitude, co$latitude, co$Al, 
       lines = copoly, 
       lines.args = list(lwd = 2, col = 'grey'),
       points = copoints, 
       points.args = list(pch = 21, bg = 'white'),
       xlim = c(-109.1, -102),
       ylim = c(36.8, 41.1))
```

The `autoimage` function generalizes the `pimage` function to 
allow for a sequence of images in a single graphic.

```{r}
# plot irregularly gridded images with separate legends
# and usa border
autoimage(lon, lat, tasmax[,,1:2], common.legend = FALSE, map = 'usa')
# with a common border and vertical scale
autoimage(lon, lat, tasmax[,,1:2], legend = "v", map = 'usa')
```
