---
title: "Heated scatterplots"
author: "Joshua P. French"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scatterplot version of heat maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## Introduction
The **autoimage** package makes it easy to plot a sequence
of images with corresponding color scales, i.e., a sequence
of heat maps, with straightforward, native options for
projection of geographical coordinates.  The package makes
it simple to add lines, points, and other features to the
images, even when the coordinates are projected.

The package also makes it easy to create a "heated 
scatterplot", which is similar to a heat map in that 
color is used to distinguish smaller and larger values of 
a variable observed over a two-dimensional domain, but is
different from a heat map in that the observed data locations
are not gridded. Instead of interpolating the data locations
onto a grid, the colors are displayed through a typical
two-dimensional scatterplot.

The `autopoints` function is the main function for creating
a sequence of heated scatterplots; it relies on the 
helper function `heat_ppoints`, which can be used in more
complex scenarios. We provide the examples below with brief
explanation.

## `autopoints` Examples
We provide examples using the non-gridded Aluminum and
Cadmium measurements for the state of Colorado, which are 
available in the **gear** package.

Here is a basic heated scatterplot for Aluminum with custom
x and y axis labels.

```{r, fig.height=5, fig.width=4}
library(autoimage)
data(co, package = "gear")
autopoints(co$longitude, co$latitude, co$Al, 
       xlab = "lon", ylab = "lat")
```

In the sets of examples, we display heated scatterplots for 
Aluminum and Cadmium while allowing them to have different z
limits and z limit breaks for the color scale. We illustrate
a few different combinations using defaults and by
specifying `zlim` and `breaks` as lists.

The following example uses the `autopoints` default options, 
which automatically determines the color scale break labels
using the the **base** `pretty` function.
```{r, fig.height=5, fig.width=8}
autopoints(co$longitude, co$latitude, co[, c("Al", "Ca")], 
       xlab = "lon", ylab = "lat", common.legend = FALSE,
       main = c("Al", "Ca"))
```

In the next example, we manually specify the z limits,
after using the `summary` function to determine appropriate
limis.
```{r, fig.height=5, fig.width=8}
summary(co[, c("Al", "Ca")])
autopoints(co$longitude, co$latitude, co[, c("Al", "Ca")], 
       xlab = "lon", ylab = "lat", common.legend = FALSE,
       main = c("Al", "Ca"), zlim = list(c(0, 10), c(0, 22)))
```

Finally, we manually specify the breaks of each variable, which 
allows for different numbers of colors for each plot.
```{r, fig.height=5, fig.width=8}
autopoints(co$longitude, co$latitude, co[, c("Al", "Ca")], 
       xlab = "lon", ylab = "lat", common.legend = FALSE,
       main = c("Al", "Ca"), 
       breaks = list(seq(0, 10, by = 2.5),
                     c(0, 2, 4, 6, 8, 22)))
```

In the following example, we project the coordinates using the 
Bonne projection. This automatically adds longitude/latitude
lines (graticules), which we lighten dramatically through
the `col` argument of the `paxes.args` list. The
graticules do not extend far enough by default, so we extend
them further through the `axis.args` via the `xat` and
`yat` arguments. We also add text above the Aluminum and
Cadmium plots via the `main` argument and a common title
via the `outer.title` argument. The `mtext.args` argument
list can be used to customize the outer title.

```{r, fig.height=5, fig.width=8}
autopoints(co$longitude, co$latitude, co[, c("Al", "Ca")], 
       xlab = "lon", ylab = "lat",
       proj = "bonne", parameters = 40,
       paxes.args = list(col = "lightgrey"),
       axis.args = list(yat = 35:43, 
                        xat = -(110:101)),
       main = c("(a) Aluminum", "(b) Cadmium"),
       outer.title = "Geochemical measurements",
       mtext.args = list(col = "blue", cex = 2))
```

We can add selected borders via the `map` arguments (which
makes use of the the `maps::map` function. We can also add
borders/lines via the lines arguments. We demnstrate using
USA county maps (`map = "county"`) and Colorado state
borders (`lines = copoly`) where `copoly` is a polygon list
available in the **autoimage** package. The lines can 
be customized through the `lines.args` argument list.

```{r, fig.height=5, fig.width=4}
autopoints(co$longitude, co$latitude, co$Al, 
       xlab = "lon", ylab = "lat", map = "county",
       lines.args = list(col = "magenta"))
```

```{r, fig.height=5, fig.width=4}
data(copoly)
autopoints(co$longitude, co$latitude, co$Al, 
       xlab = "lon", ylab = "lat", lines = copoly,
       lines.args = list(col = "orange", lwd = 3))
```

Naturally, one may wish to change the color palette or the
breaks associated with the colors. This can be done via
the `col` and `breaks` arguments. `breaks` must be a sequential
vector with length one element longer than `col`. We also
move the legend scale to the right side.

```{r, fig.height=4, fig.width=6}
autopoints(co$longitude, co$latitude, co$Al, legend = "v",
           xlab = "lon", ylab = "lat",
           col = colorspace::sequential_hcl(n = 4, palette = "Plasma"),
           breaks = c(0, 1, 2, 3, 8))
```

We add the locations and names of two Colorado cities to the
Colorado geochemical data using the `points` and `text` 
arguments. The appearance of these can be customized through
the `points.args` and `text.args` argument lists,
respectively.

```{r, fig.height=4, fig.width=7}
citypoints = list(x = c(-104.98, -104.80), y = c(39.74, 38.85),
                  labels = c("Denver", "Colorado Springs"))
autopoints(co$lon, co$lat, co[,c("Al", "Ca")], common.legend = FALSE, 
           main = c("Aluminum", "Cadmium"), 
           points = citypoints,
           points.args = list(pch = 6, col = "magenta"),
           text = citypoints,
           text.args = list(pos = 3, col = "orange"),
           xlab = "lon", ylab = "lat")
```

## Richer plots using `autolayout` and `autolegend`

Suppose we want to add custom features to a sequences of images and heated scatterplots, with each plots receiving different features.  One can create a richer sequence of images using the `autolayout` and `autolegend` functions.

The `autolayout` function partitions the graphic device into the sections needed to create a sequence of images.  The most important function arguments include `size`, `legend`, `common.legend`, and `lratio`, which correspond to the same arguments in the `autoimage` function.  The `outer` argument specifies whether an `outer.title` is desired.  The default is `FALSE`.  By default, numbers identify the plotting order of the sections, though these can be hidden by setting `show = FALSE`.  As an initial example, we create a 2 $\times$ 3 grid of images with a common vertical legend.

We now create a complicated (though unrealistic) example of this.  We first extract the borders of Hawaii and Alaska from the `"world"` map in the **maps** package and store it as the `hiak` list.  We then select a small subset of cities in Colorado from the `us.cities` dataset in the **maps** package and store this in the `codf` data frame.  Lastly, we select the U.S. capitals from the `us.cities` dataset and store this in the `capdf` data frame.  

```{r}
# load world map
data(worldMapEnv, package = "maps")
# extract hawaii and alaskan borders
hiak <- maps::map("world", c("USA:Hawaii", "USA:Alaska"), 
                  plot = FALSE)
# load us city information
data(us.cities, package = "maps")
# extract colorado cities from us.cities
codf <- us.cities[us.cities$country.etc == "CO", ]
# select smaller subset of colorado cities
# extract capitals from us.cities
capdf <- us.cities[us.cities$capital == 2,]
```

Having obtained the relevant information, we setup a 1 $\times$ 2 matrix of images with individual horizontal legends and an area for a common title.  We create an image plot of NARCCAP data using the mercator projection and including grey state borders.  The borders of Hawaii and Alaska are added using the `plines` function.  The state capitals are added to the image using the `ppoints` function.  The first image is then titled using the `title` function.  The legend is then added using the `autolegend` function.  Next, a heated scatterplots of the Colorado Aluminum measurements is created using the `heat_ppoints` function.  The coordinates are projected using the Bonne projection, the color scheme is customized, grey county borders are added to the plot, but the grid lines are removed.  The `ppoints` function is then used to add locations for several Colorado cities to the image.  The `ptext` function is then used to add the names of these cities to the image.  The second heated scatterplot is then titled using the `title` function. The appropriate legend is then added using the `autolegend` function.  Lastly, a common title is added using the `mtext` function.  

```{r, fig.width=7, fig.height=5, hold=TRUE}
# setup plotting area
autolayout(c(1, 2), legend = "h", common.legend = FALSE, outer = TRUE)
# create image of NARCCAP data.
# xlim is chosen so to include alaska and hawaii
# add grey state borders
# extend graticules (longitude/latitude grid lines)
pimage(lon, lat, tasmax[,,1], legend = "none",
       proj = "mercator", map = "state",
       xlim = c(-180, 20),
       axis.args = list(xat = seq(-175, -25, by = 25),
                        yat = seq(-10, 80, len = 10)),
       lines.args = list(col = "grey"))
# add hawaii and alaskan borders
plines(hiak, proj = "mercator", col = "grey")
# add state captials to image
ppoints(capdf$lon, capdf$lat, proj = "mercator", pch = 16)
# title image
title("tasmax for North America")
# add legend for plot
autolegend()
# load colorado geochemical data
data(co, package = "gear")
# create image for colorado aluminum measurements
# use bonne projection
# customize legend colors
# add grey county borders
# exclude longitude/latitude
heat_ppoints(co$lon, co$lat, co$Al, map = "county", legend = "none",
       proj = "bonne", parameters = 39, 
       paxes.args = list(grid = FALSE),
       col = cm.colors(5),
       lines.args = list(col = "grey"),
       xlab = "lon", ylab = "lat")
# add colorado city points to image
ppoints(codf$lon, codf$lat, pch = 16, proj = "bonne")
# add names of colorado cities to image
ptext(codf$lon, codf$lat, labels = codf$name, proj = "bonne", pos = 4)
# title plot
title("Colorado Aluminum levels (%)")
# add legend to current image
autolegend()
# add common title for plots
mtext("Two complicated maps", col = "purple", outer = TRUE, cex = 2)
```
